---
title: "Assignment 1 – Task 2: MR-QAP Regression"
subtitle: "Network Modeling SS 2026"
author: "Group XX"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# Setup

```{r libraries}
library(sna)
library(network)

set.seed(161)

# Path to the cleaned adjacency matrix CSVs (relative to this file)
data_dir  <- "cleaned_csv"
# Attribute file (comma-separated, produced by cleaning.R via write.csv)
attr_file <- file.path(data_dir, "cleaned_attr.csv")

classrooms <- c("01", "02", "03", "04", "05", "09", "10", "12")
```

## Helper functions

```{r helpers}
# Load an adjacency matrix from a cleaned CSV.
# The cleaning pipeline writes semicolon-separated files with row names.
# Classrooms 01-09 developed a leading-zero inconsistency during cleaning
# (row names lost the leading zero, column names kept it).
# We normalise both to the no-leading-zero form so that
# intersect(rownames, colnames) works correctly across all classrooms.
load_adj <- function(file) {
  mat_raw <- read.csv(file, sep = ";", row.names = 1,
                      header = TRUE, check.names = FALSE)
  rownames(mat_raw) <- sub("^0+", "", rownames(mat_raw))
  colnames(mat_raw) <- sub("^0+", "", colnames(mat_raw))
  as.matrix(mat_raw)
}

# Build cleaned predictor matrices for a given classroom ID string
# (e.g. "10" -> cleaned_10_W1.csv / cleaned_10_W2.csv).
build_data <- function(classroom) {
  f1 <- file.path(data_dir, paste0("cleaned_", classroom, "_W1.csv"))
  f2 <- file.path(data_dir, paste0("cleaned_", classroom, "_W2.csv"))

  W1_raw <- load_adj(f1)
  W2_raw <- load_adj(f2)

  # Restrict to students present in both waves
  ids_common <- intersect(rownames(W1_raw), rownames(W2_raw))
  W1 <- W1_raw[ids_common, ids_common]
  W2 <- W2_raw[ids_common, ids_common]
  storage.mode(W1) <- "numeric"
  storage.mode(W2) <- "numeric"

  # Flag nodes with any NA in their nomination row in either wave
  # (partial non-response; all-NA rows were already removed by cleaning.R)
  missing_net <- rowSums(is.na(W1)) > 0 | rowSums(is.na(W2)) > 0

  # Match attributes; students absent from attr_raw get an all-NA row
  attr_sub <- attr_raw[match(ids_common, as.character(attr_raw$studentID)), ]

  # Flag nodes with any missing required covariate
  missing_attr <- is.na(attr_sub$literacy_end) |
                  is.na(attr_sub$HISEI) |
                  is.na(attr_sub$gender)

  keep       <- !(missing_net | missing_attr)
  ids_clean  <- ids_common[keep]
  W1_c       <- W1[keep, keep]
  W2_c       <- W2[keep, keep]
  attr_c     <- attr_sub[keep, ]

  n   <- nrow(W1_c)
  lit <- as.numeric(attr_c$literacy_end)
  gen <- attr_c$gender
  his <- as.numeric(attr_c$HISEI)

  # Predictor matrix: receiver literacy  X[i,j] = literacy_end of j
  X_reclit <- matrix(rep(lit, each = n), nrow = n, ncol = n)

  # Predictor matrix: same-gender dyad  X[i,j] = 1 if gender_i == gender_j
  X_gender <- outer(gen, gen, FUN = function(a, b) as.integer(a == b))

  # Predictor matrix: sender HISEI  X[i,j] = HISEI of i
  X_hisei <- matrix(rep(his, times = n), nrow = n, ncol = n)

  # Zero diagonals (no self-nominations)
  diag(W1_c) <- diag(W2_c) <- 0
  diag(X_reclit) <- diag(X_gender) <- diag(X_hisei) <- 0

  list(Y = W2_c, X_w1 = W1_c, X_reclit = X_reclit,
       X_gender = X_gender, X_hisei = X_hisei,
       ids = ids_clean, n = n)
}

# Run MR-QAP regression (netlm with QAP permutation)
run_qap <- function(dat, reps = 5000) {
  netlm(dat$Y,
        list(dat$X_w1, dat$X_reclit, dat$X_gender, dat$X_hisei),
        reps    = reps,
        mode    = "digraph",
        nullhyp = "qap")
}

# Extract a tidy result data-frame from a netlm fit
tidy_netlm <- function(fit, classroom) {
  data.frame(
    classroom = classroom,
    term      = c("Intercept", "W1_lag", "recv_literacy",
                  "same_gender", "send_HISEI"),
    estimate  = round(fit$coefficients, 4),
    p_value   = round(fit$pgreqabs,     4),
    row.names = NULL
  )
}
```

## Load attribute data

```{r load-attr}
# cleaned_attr.csv: comma-separated, produced by cleaning.R (na.omit applied,
# so only students with complete literacy_end, HISEI, and gender are kept).
attr_raw <- read.csv(attr_file, sep = ",", stringsAsFactors = FALSE)
attr_raw$studentID <- as.character(attr_raw$studentID)
```

---

# Task 2.1 – Simple QAP for Classroom 10

We first assess whether friendship nominations in Wave 2 are predicted by
the same nominations in Wave 1, using a simple QAP regression with the
lagged network as the only predictor:

$$Y_{ij}^{W2} = \beta_0 + \beta_1 \cdot Y_{ij}^{W1} + \varepsilon_{ij}$$

The **null hypothesis** is that the observed association between $W1$ and $W2$
could arise by chance under random permutation of the network.

```{r simple-qap-10}
dat10 <- build_data("10")
cat("Classroom 10 – nodes retained after cleaning:", dat10$n, "\n")

fit10_simple <- netlm(dat10$Y, list(dat10$X_w1),
                      reps = 5000, mode = "digraph", nullhyp = "qap")
summary(fit10_simple)
```

**Interpretation:** The coefficient on the lagged Wave-1 network is positive
and statistically significant ($p < 0.05$), indicating that friendship ties
present at Wave 1 are strongly predictive of ties at Wave 2.  This is
consistent with *friendship stability* (tie persistence / network inertia):
students who were friends at the beginning of the school year largely remain
friends by the end.

---

# Task 2.2 – Operationalisation of Hypotheses

The full MR-QAP model is:

$$Y_{ij}^{W2} = \beta_0
  + \beta_1 Y_{ij}^{W1}
  + \beta_2 X_{ij}^{\text{reclit}}
  + \beta_3 X_{ij}^{\text{gender}}
  + \beta_4 X_{ij}^{\text{HISEI}}
  + \varepsilon_{ij}$$

We add three dyadic predictor matrices to test the following hypotheses:

| Hypothesis | Operationalisation | Entry $X_{ij}$ |
|---|---|---|
| i. High-literacy receivers get *fewer* nominations | **Receiver literacy** | $\text{literacy\_end}_j$ |
| ii. Same-gender pairs more likely to nominate | **Same gender** | $\mathbf{1}[\text{gender}_i = \text{gender}_j]$ |
| iii. High-HISEI senders nominate *more* | **Sender HISEI** | $\text{HISEI}_i$ |

**Why these operationalisations?**

- *Hypothesis i* is an **alter (receiver) effect**: variation in how often a
  node is *chosen* is captured by a column-constant matrix whose entry in
  column $j$ equals $j$'s literacy score.
- *Hypothesis ii* is a **homophily / dyadic covariate**: a binary indicator
  equals 1 when sender and receiver share the same gender.
- *Hypothesis iii* is an **ego (sender) effect**: variation in how many
  nominations a node *sends* is captured by a row-constant matrix whose entry
  in row $i$ equals $i$'s HISEI score.

---

# Task 2.3 – MR-QAP Estimation for Classroom 10

```{r full-qap-10}
fit10 <- run_qap(dat10, reps = 5000)
summary(fit10)
```

```{r tidy-10}
tidy10 <- tidy_netlm(fit10, "10")
knitr::kable(tidy10, digits = 4,
             caption = "MR-QAP coefficients – Classroom 10")
```

## Interpretation

The QAP permutation test randomly shuffles row and column labels of the
response matrix to obtain the null distribution of each coefficient; the
reported $p$-value (`pgreqabs`) is the proportion of permuted coefficients at
least as extreme (in absolute value) as the observed one.

| Term | Expected direction | Observed | Supported? |
|---|---|---|---|
| `W1_lag` | positive (stability) | `r round(tidy10$estimate[tidy10$term=="W1_lag"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="W1_lag"],3)`) | Yes |
| `recv_literacy` | negative (H i) | `r round(tidy10$estimate[tidy10$term=="recv_literacy"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="recv_literacy"],3)`) | See text |
| `same_gender` | positive (H ii) | `r round(tidy10$estimate[tidy10$term=="same_gender"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="same_gender"],3)`) | See text |
| `send_HISEI` | positive (H iii) | `r round(tidy10$estimate[tidy10$term=="send_HISEI"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="send_HISEI"],3)`) | See text |

- **Lagged network (W1):** The large, highly significant positive coefficient
  confirms strong **friendship stability**.  A prior tie is by far the
  strongest predictor of a future tie.

- **Receiver literacy (H i):** A negative coefficient would indicate that
  high-literacy students attract fewer nominations; a positive one would
  indicate the opposite.  The sign and permutation $p$-value determine
  whether this hypothesis is supported in Classroom 10.

- **Same gender (H ii):** A positive, significant coefficient supports the
  **gender-homophily** hypothesis: students preferentially nominate same-gender
  peers.

- **Sender HISEI (H iii):** A positive, significant coefficient would indicate
  that students from higher socioeconomic backgrounds send more friendship
  nominations.

---

# Task 2.4 – All Remaining Classrooms

```{r all-classrooms, results='asis', cache=TRUE}
all_results <- list()
all_tidy    <- list()

for (cl in classrooms) {
  dat <- tryCatch(build_data(cl), error = function(e) NULL)
  if (is.null(dat) || dat$n < 5) next

  fit              <- run_qap(dat, reps = 5000)
  all_results[[cl]] <- fit
  all_tidy[[cl]]    <- tidy_netlm(fit, cl)

  cat(sprintf("\n## Classroom %s (n = %d)\n\n", cl, dat$n))
  print(knitr::kable(all_tidy[[cl]], digits = 4,
                     caption = sprintf("Classroom %s", cl)))
  cat("\n")
}

all_tidy_df <- do.call(rbind, all_tidy)
```

---

# Task 2.5 – Cross-Classroom Summary and Generalisation

## Combined coefficient table

```{r combined-table}
wide <- reshape(all_tidy_df,
                idvar     = "classroom",
                timevar   = "term",
                direction = "wide")
colnames(wide) <- gsub("estimate\\.", "", colnames(wide))
colnames(wide) <- gsub("p_value\\.", "p_",  colnames(wide))
knitr::kable(wide, digits = 3,
             caption = "Coefficients and permutation p-values across all classrooms")
```

## Caterpillar plots

Each panel shows the estimated coefficient for one predictor across all
classrooms, sorted by effect size.  **Blue** points have $p < 0.05$;
**grey** points are not statistically significant.  The dashed red line
marks zero (no effect).

```{r caterpillar, fig.height=8, fig.width=10}
terms_plot <- c("W1_lag", "recv_literacy", "same_gender", "send_HISEI")
term_labels <- c(
  W1_lag        = "Lagged friendship (W1)",
  recv_literacy = "Receiver literacy",
  same_gender   = "Same gender",
  send_HISEI    = "Sender HISEI"
)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (trm in terms_plot) {
  sub  <- all_tidy_df[all_tidy_df$term == trm, ]
  sub  <- sub[order(sub$estimate), ]
  cols <- ifelse(sub$p_value < 0.05, "steelblue", "grey60")

  rng <- range(sub$estimate)
  pad <- diff(rng) * 0.15

  plot(sub$estimate, seq_len(nrow(sub)),
       xlab  = "Coefficient estimate",
       ylab  = "Classroom",
       main  = term_labels[trm],
       pch   = 19,
       col   = cols,
       yaxt  = "n",
       xlim  = c(rng[1] - pad, rng[2] + pad))
  axis(2, at = seq_len(nrow(sub)), labels = sub$classroom, las = 1)
  abline(v = 0, lty = 2, col = "red")
  legend("topright",
         legend = c("p < 0.05", "p \u2265 0.05"),
         col    = c("steelblue", "grey60"),
         pch    = 19, cex = 0.75, bty = "n")
}
```

## Significance count by predictor

```{r sign-summary}
sig_counts <- tapply(all_tidy_df$p_value < 0.05,
                     all_tidy_df$term, sum)
print(sig_counts)
```

## Discussion

**Friendship stability (`W1_lag`):** The lagged wave-1 network has a positive
and statistically significant coefficient in nearly every classroom.  This is
the most robust finding: friendships formed at the start of the school year
largely persist until the end, consistent with what the literature describes
as *tie persistence* or *network inertia*.

**Receiver literacy (Hypothesis i):** The direction and significance of this
coefficient vary across classrooms, suggesting the relationship between
academic achievement and popularity is not uniform.  In some classrooms a
negative coefficient (fewer nominations for high-literacy students) appears;
in others the effect is negligible or reversed.  Hypothesis i therefore
receives only *weak and inconsistent* support.

**Same gender (Hypothesis ii):** Gender homophily—the tendency to nominate
same-gender peers more often—is present in several classrooms with a positive
coefficient.  However, the effect is not universally significant, indicating
that gender segregation in peer nominations varies in strength across
classroom contexts.

**Sender HISEI (Hypothesis iii):** The association between socioeconomic
status and outgoing friendship nominations is modest and inconsistent.
High-HISEI students do not systematically nominate more friends in every
setting, suggesting that this effect, if present, depends on classroom
composition and context.

**Generalisability:** The findings rest on only eight classrooms from a
single Czech school cohort, limiting external validity.  The heterogeneity
in effect sizes and significance levels across classrooms underscores that
adolescent peer-network dynamics are context-dependent.  Classroom
composition (gender ratio, HISEI distribution, class size) likely moderates
individual effects.  A broader, multi-school sample would be required for
generalisation.  Differences across classrooms for the gender and literacy
predictors could reflect real variation in how these characteristics shape
friendship selection, or they could reflect sampling variability given the
relatively small number of students per classroom.
