---
title: "Assignment 1 – Task 2: MR-QAP Regression"
subtitle: "Network Modeling SS 2026"
author: "Group XX"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

# Setup

```{r libraries}
library(sna)
library(network)

set.seed(161)

# Path to the cleaned adjacency matrix CSVs (relative to this file)
data_dir  <- "cleaned_csv"
# Attribute file (comma-separated, produced by cleaning.R via write.csv)
attr_file <- file.path(data_dir, "cleaned_attr.csv")

classrooms <- c("01", "02", "03", "04", "05", "09", "10", "12")
```

## Helper functions

We begin by defining some helper functions. `build_data` is our cleaning pipeline, 
`run_qap` is a cleaner expression of the `netlm` function, and `tidy_netlm` gets a tidy 
dataframe from the `netlm` fit. 

```{r helpers}
# Load an adjacency matrix from a cleaned CSV.
# The cleaning pipeline writes semicolon-separated files with row names.
load_adj <- function(file) {
  mat_raw <- read.csv(file, sep = ";", row.names = 1,
                      header = TRUE, check.names = FALSE)
  rownames(mat_raw) <- sub("^0+", "", rownames(mat_raw))
  colnames(mat_raw) <- sub("^0+", "", colnames(mat_raw))
  as.matrix(mat_raw)
}

# Build cleaned predictor matrices for a given classroom ID string
# (e.g. "10" -> cleaned_10_W1.csv / cleaned_10_W2.csv).
build_data <- function(classroom) {
  f1 <- file.path(data_dir, paste0("cleaned_", classroom, "_W1.csv"))
  f2 <- file.path(data_dir, paste0("cleaned_", classroom, "_W2.csv"))

  W1_raw <- load_adj(f1)
  W2_raw <- load_adj(f2)

  # Restrict to students present in both waves
  ids_common <- intersect(rownames(W1_raw), rownames(W2_raw))
  W1 <- W1_raw[ids_common, ids_common]
  W2 <- W2_raw[ids_common, ids_common]
  storage.mode(W1) <- "numeric"
  storage.mode(W2) <- "numeric"

  # Flag nodes with any NA in their nomination row in either wave
  # (partial non-response; all-NA rows were already removed by cleaning.R)
  missing_net <- rowSums(is.na(W1)) > 0 | rowSums(is.na(W2)) > 0

  # Match attributes; students absent from attr_raw get an all-NA row
  attr_sub <- attr_raw[match(ids_common, as.character(attr_raw$studentID)), ]

  # Flag nodes with any missing required covariate
  missing_attr <- is.na(attr_sub$literacy_end) |
                  is.na(attr_sub$HISEI) |
                  is.na(attr_sub$gender)

  keep       <- !(missing_net | missing_attr)
  ids_clean  <- ids_common[keep]
  W1_c       <- W1[keep, keep]
  W2_c       <- W2[keep, keep]
  attr_c     <- attr_sub[keep, ]

  n   <- nrow(W1_c)
  lit <- as.numeric(attr_c$literacy_end)
  gen <- attr_c$gender
  his <- as.numeric(attr_c$HISEI)

  # Predictor matrix: receiver literacy  X[i,j] = literacy_end of j
  X_reclit <- matrix(rep(lit, each = n), nrow = n, ncol = n)

  # Predictor matrix: same-gender dyad  X[i,j] = 1 if gender_i == gender_j
  X_gender <- outer(gen, gen, FUN = function(a, b) as.integer(a == b))

  # Predictor matrix: sender HISEI  X[i,j] = HISEI of i
  X_hisei <- matrix(rep(his, times = n), nrow = n, ncol = n)

  # Zero diagonals (no self-nominations)
  diag(W1_c) <- diag(W2_c) <- 0
  diag(X_reclit) <- diag(X_gender) <- diag(X_hisei) <- 0

  list(Y = W2_c, X_w1 = W1_c, X_reclit = X_reclit,
       X_gender = X_gender, X_hisei = X_hisei,
       ids = ids_clean, n = n)
}

# Run MR-QAP regression (netlm with QAP permutation)
run_qap <- function(dat, reps = 5000) {
  netlm(dat$Y,
        list(dat$X_w1, dat$X_reclit, dat$X_gender, dat$X_hisei),
        reps    = reps,
        mode    = "digraph",
        nullhyp = "qap")
}

# Extract a tidy result data-frame from a netlm fit
tidy_netlm <- function(fit, classroom) {
  data.frame(
    classroom = classroom,
    term      = c("Intercept", "W1_lag", "recv_literacy",
                  "same_gender", "send_HISEI"),
    estimate  = round(fit$coefficients, 4),
    p_value   = round(fit$pgreqabs,     4),
    row.names = NULL
  )
}
```

## Load attribute data

```{r load-attr}
attr_raw <- read.csv(attr_file, sep = ",", stringsAsFactors = FALSE)
attr_raw$studentID <- as.character(attr_raw$studentID)
```

---

# Task 2.1 – Simple QAP for Classroom 10

We first assess whether friendship nominations in Wave 2 are predicted by
the same nominations in Wave 1, using a simple QAP regression with the
lagged network as the only predictor:

$$Y_{ij}^{W2} = \beta_0 + \beta_1 \cdot Y_{ij}^{W1} + \varepsilon_{ij}$$

The **null hypothesis** is that the observed association between $W1$ and $W2$
is weak enough such that it could reasonably arise by chance under random permutation of the network.

```{r simple-qap-10}
dat10 <- build_data("10")
cat("Classroom 10 – nodes retained after cleaning:", dat10$n, "\n")

fit10_simple <- netlm(dat10$Y, list(dat10$X_w1),
                      reps = 5000, mode = "digraph", nullhyp = "qap")
summary(fit10_simple)
```

**Interpretation:** The coefficient on the lagged Wave-1 network is positive ($\beta_1 \approx 0.52$)
and significant at the 5% level, which means we can reject the null. 
Friendships present at Wave 1 are strongly predictive of ties at Wave 2. This is what we'd 
expect to see intuitively: friendships are stable across time (relatively, at least--this is high school, after all). In network science terminology, 
we could say that our network has some intertia or tie-persistence. 

---

# Task 2.2 – Operationalisation of Hypotheses

The full MR-QAP model is:

$$Y_{ij}^{W2} = \beta_0
  + \beta_1 Y_{ij}^{W1}
  + \beta_2 X_{ij}^{\text{reclit}}
  + \beta_3 X_{ij}^{\text{gender}}
  + \beta_4 X_{ij}^{\text{HISEI}}
  + \varepsilon_{ij}$$

We add three dyadic predictor matrices to test the following hypotheses:

| Hypothesis | Operationalisation | Entry $X_{ij}$ |
|---|---|---|
| i. High-literacy receivers get *fewer* nominations | **Receiver literacy** | $\text{literacy\_end}_j$ |
| ii. Same-gender pairs more likely to nominate | **Same gender** | $\mathbf{1}[\text{gender}_i = \text{gender}_j]$ |
| iii. High-HISEI senders nominate *more* | **Sender HISEI** | $\text{HISEI}_i$ |

We tried to choose the simplest possible representation of each effect. 

- *Hypothesis i* is an **alter (receiver) effect**: variation in how often a
  node is *chosen* is captured by a column-constant matrix whose entry in
  column $j$ equals $j$'s literacy score.
- *Hypothesis ii* is a **homophily / dyadic covariate**: a binary indicator
  equals 1 when sender and receiver share the same gender. No other encoding would make sense here. 
- *Hypothesis iii* is an **ego (sender) effect**: variation in how many
  nominations a node *sends* is captured by a row-constant matrix whose entry
  in row $i$ equals $i$'s HISEI score.

---

# Task 2.3 – MR-QAP Estimation for Classroom 10

We re-run the model...

```{r full-qap-10}
fit10 <- run_qap(dat10, reps = 5000)
summary(fit10)
```

...and obtain:

```{r tidy-10}
tidy10 <- tidy_netlm(fit10, "10")
knitr::kable(tidy10, digits = 4,
             caption = "MR-QAP coefficients – Classroom 10")
```

## Interpretation

The QAP permutation test randomly shuffles row and column labels of the
response matrix to obtain the null distribution of each coefficient. The
reported $p$-value (`pgreqabs`) is the proportion of permuted coefficients at
least as extreme (in absolute value) as the observed one.

| Term | Expected direction | Observed | Supported? |
|---|---|---|---|
| `W1_lag` | positive (stability) | `r round(tidy10$estimate[tidy10$term=="W1_lag"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="W1_lag"],3)`) | Yes |
| `recv_literacy` | negative | `r round(tidy10$estimate[tidy10$term=="recv_literacy"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="recv_literacy"],3)`) | No |
| `same_gender` | positive | `r round(tidy10$estimate[tidy10$term=="same_gender"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="same_gender"],3)`) | Yes |
| `send_HISEI` | positive | `r round(tidy10$estimate[tidy10$term=="send_HISEI"],3)` ($p$ = `r round(tidy10$p_value[tidy10$term=="send_HISEI"],3)`) | No |

We now comment on each hypothesis: 

- *Lagged network (W1):* Even when controlling for three further variables, 
 we find that the tie persistence property is well-supported at the 5%-level.
 In fact, of the correlates investigated, being friends in a prior period is the 
 strongest determinant of current friendship together with gender (the effect strength is equal).

- *Receiver literacy.* We were asked to test the one-sided hypothesis that literacy scores 
  negatively correlate with friendship nominations. That means we would have expected a negative coefficient. 
  We found no evidence of this: the parameter estimate was close to zero and not supported at the 5% level. We fail to reject the null. 

- *Same gender.* A positive, significant coefficient supports the
  gender-homophily hypothesis: students prefer being friends with same-gender
  peers. Interestingly, the effect strength (and significance) is very close to the tie persistence hypothesis. 

- *Sender HISEI.* A positive, significant coefficient would indicate
  that students from higher socioeconomic backgrounds tend to send more friendship
  nominations. As with the literacy hypothesis, we find no conclusive information. The 
  estimate is close to zero and the significance is far too low to reject at the 5% level. (In fact, 
  a p-value of 0.18 indicates that we'd expect to see such an effect in every fifth simulated network.) 

---

# Task 2.4 – All Remaining Classrooms

We re-run our analysis and provide output tables. In broad strokes, our findings remain broadly the same. There are some aberrations--we comment on these in 6.4. 

```{r all-classrooms, results='asis', cache=TRUE}
all_results <- list()
all_tidy    <- list()

for (cl in classrooms) {
  dat <- tryCatch(build_data(cl), error = function(e) NULL)
  if (is.null(dat) || dat$n < 5) next

  fit              <- run_qap(dat, reps = 5000)
  all_results[[cl]] <- fit
  all_tidy[[cl]]    <- tidy_netlm(fit, cl)

  cat(sprintf("\n## Classroom %s (n = %d)\n\n", cl, dat$n))
  print(knitr::kable(all_tidy[[cl]], digits = 4,
                     caption = sprintf("Classroom %s", cl)))
  cat("\n")
}

all_tidy_df <- do.call(rbind, all_tidy)
```

---

# Task 2.5 – Cross-Classroom Summary and Generalisation

Overall, we have found the following significant effects: 

```{r sign-summary}
sig_counts <- tapply(all_tidy_df$p_value < 0.05,
                     all_tidy_df$term, sum)
print(sig_counts)
```


## Caterpillar plots

We now visualize our data via caterpillar plots. 
Each panel shows the estimated coefficient for one predictor across all
classrooms, sorted by effect size.  Blue points are significant at the $p < 0.05$ level. 

```{r caterpillar, fig.height=8, fig.width=10}
terms_plot <- c("W1_lag", "recv_literacy", "same_gender", "send_HISEI")
term_labels <- c(
  W1_lag        = "Lagged friendship (W1)",
  recv_literacy = "Receiver literacy",
  same_gender   = "Same gender",
  send_HISEI    = "Sender HISEI"
)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
for (trm in terms_plot) {
  sub  <- all_tidy_df[all_tidy_df$term == trm, ]
  sub  <- sub[order(sub$estimate), ]
  cols <- ifelse(sub$p_value < 0.05, "steelblue", "grey60")

  rng <- range(sub$estimate)
  pad <- diff(rng) * 0.15

  plot(sub$estimate, seq_len(nrow(sub)),
       xlab  = "Coefficient estimate",
       ylab  = "Classroom",
       main  = term_labels[trm],
       pch   = 19,
       col   = cols,
       yaxt  = "n",
       xlim  = c(rng[1] - pad, rng[2] + pad))
  axis(2, at = seq_len(nrow(sub)), labels = sub$classroom, las = 1)
  abline(v = 0, lty = 2, col = "red")
  legend("topright",
         legend = c("p < 0.05", "p \u2265 0.05"),
         col    = c("steelblue", "grey60"),
         pch    = 19, cex = 0.75, bty = "n")
}
```

## Discussion

**Friendship stability (`W1_lag`):** The lagged wave-1 network has a positive
and statistically significant coefficient in nearly every classroom.  This is
our most robust finding: friendships formed at the start of the school year
largely persist until the end, consistent with what the literature describes
as *tie persistence* or *network inertia*.

**Receiver literacy (Hypothesis i):** The direction and significance of this
coefficient varies across classrooms, suggesting the relationship between
academic achievement and popularity is not uniform.  In some classrooms a
negative coefficient (fewer nominations for high-literacy students) appears;
in others the effect is negligible or reversed.  Hypothesis i therefore
receives only *weak and inconsistent* support.

**Same gender (Hypothesis ii):** Gender homophily—the tendency to nominate
same-gender peers more often—is present in several classrooms with a positive
coefficient.  However, the effect is not universally significant (e.g. classrooms 2 and 14). 
The reasons for this would merit further research. Are there other variables that describe 
when (and, if we took a causal view, why) gender homophily breaks down? 

**Sender HISEI (Hypothesis iii):** The association between socioeconomic
status and outgoing friendship nominations is modest and inconsistent.
High-HISEI students do not systematically nominate more friends in every
setting, suggesting that this effect, if present, depends on classroom
composition and context. We do find that classroom 9's socioeconomic hypothesis 
is (just barely) significant at the 5% level. We are cautious and chalk this up to multiple testing: a 5% significance means that we'd expect a significant 
effect for every 20 hypotheses by pure chance. We have tested far more than 20. Further, the coefficient value ($\approx 0.0021$) is hardly inspiring. 

**Generalisability:** The findings rest on only eight classrooms from a
single Czech school cohort, limiting external validity.  The heterogeneity
in effect sizes and significance levels across classrooms underscores that
adolescent peer-network dynamics are context-dependent.  Classroom
composition (gender ratio, HISEI distribution, class size) likely moderates
individual effects.  A broader, multi-school sample would be required for
generalisation.  Differences across classrooms for the gender and literacy
predictors could reflect real variation in how these characteristics shape
friendship selection, or they could reflect sampling variability given the
relatively small number of students per classroom.